Cognate beginnings to bilingual lexical acquisition

### 2023-06-27 manuscript/manuscript.qmd
Fix figure and add new comments

The foundations of word learning are in place early in age. At six months of age, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and, shortly after, caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar amount of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

One important observation of studies on bilinguals’ early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of cat is *gato*, a cognate word; the translation equivalent of dog is *perro*, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historic reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@@

A second set of studies concluded that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of participant’s vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. @mitchell2022cognates addressed this issue in a larger sample longitudinal study. These authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research provide a strong support to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of tree), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of table), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

Although there is converging evidence on the role of parallel activation, the previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words, and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, only if such translation equivalent is co-activated. The probability of this co-activation is proportional to the phonological similarity between the two translation equivalents, and given that cognates share higher phonological similarity than non-cognates, the former may be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across both languages: words from the lower-exposure language would receive additional learning instances through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language, roughly following a one-to-one mapping (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for dog) (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items. Each subset contained a random but representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

For each word included in the questionnaire, we manually generated a broad phonological transcription in X-SAMPA format [@wells1995computercoding]. Catalan word-forms were transcribed to Central Catalan phonology, and Spanish word-forms were transcribed to Castilian Spanish phonology. For each translation equivalent, we calculated the phonological similarity between the Catalan and the Spanish word-form using the same Levenshtein similarity measure used in @floccia2018introduction [@heeringa2003norwegian; @schepens2012distributions; see also @fourtassi2020growth; @laing2022phonological for a similar approach].

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants) for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (composite measure of Zipf-transformed lexical frequency and language degree of exposure), and *Cognateness* (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data (see @eq-linear) [@barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

@@

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome]. We implemented the model using was done using the `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four MCMC  chains with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

The main effect of $\text{Age}$ showed the strongest association with the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_age_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_age_std"]]$.lower, 2)`, `r round(post_draws_list[["b_age_std"]]$.upper, 2)`]), with all of its posterior samples falling out of the ROPE. A one-month increment in age increased a maximum of `r round(post_draws_list[["b_age_std"]]$.median/sd(responses$age), 2)` the probability of acquisition. Similarly, the word exposure index ($\text{Exposure}$) had a strong effect on the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_exposure_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std"]]$.lower, 2)`, `r round(post_draws_list[["b_exposure_std"]]$.upper, 2)`]). All of the posterior samples of this regression coefficient excluded the ROPE. The impact of this predictor on the probability of acquisition was positive: for every standard deviation increase in exposure, the participant was `r round(post_draws_list[["b_exposure_std"]]$.median/sd(responses$exposure), 2)` more likely to acquire it. Word-form length also showed a significant association with probability of acquisition ($\beta$ = `r round(post_draws_list[["b_n_phon_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_n_phon_std"]]$.lower, 2)`, `r round(post_draws_list[["b_n_phon_std"]]$.upper, 2)`]). For every phoneme in the word-form, participants were `r round((post_draws_list[["b_n_phon_std"]]$.median)/sd(responses$n_phon), 2)` less likely to know it. The 95% HDI of the regression coefficient of the $\text{Age} \times \text{Exposure}$ interaction also excluded the ROPE ($\beta$ = `r round(post_draws_list[["b_age_std:exposure_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_age_std:exposure_std"]]$.lower, 2)`, `r round(post_draws_list[["b_age_std:exposure_std"]]$.upper, 2)`]), showing that the effect of the word exposure index differed across ages: older children were more sensitive likely to acquire words with higher exposure rate than younger children.

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Lines correspond to 100 posterior-predicted medians. Different colours indicate different levels of cognateness (phonological similarity). Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are shown on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d",
                  "#003f5c",
                  "#122A36", 
                  "#ffa600", 
                  "#ff6361", 
                  "#bc5090")

model_epreds |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, 
               colour = lv, 
               fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               linetype = "dotted") +
    geom_hline(yintercept = 0.5,
               linetype = "dotted") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 2/4,
              alpha = 1/5) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```

>>>>

The foundations of word learning are in place early in age. At six months of age, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and, shortly after, caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

One important observation of studies on bilinguals' early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of *cat* is *gato*, a cognate word; the translation equivalent of *dog* is *perro*, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historic reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@@

A second set of studies concluded that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of participant’s vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual child produced cognates earlier than non-cognates, on average. @mitchell2022cognates addressed this issue in a larger sample longitudinal study. These authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research provide a strong support to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

Although there is a consensus on the role of parallel activation in bilinguals' lexical processing and acquisition, previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. The probability of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive additional activation through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language, roughly following a one-to-one mapping^[Although for some of the included words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the Catalan and Spanish lists]. (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*) (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic/functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

We manually generated a broad phonological transcription of every word included in the vocabulary checklists in X-SAMPA format [@wells1995computercoding]. Catalan word-forms were transcribed to Central Catalan phonology, and Spanish word-forms were transcribed to Castilian Spanish phonology. For each translation equivalent, we calculated the phonological similarity between the Catalan and the Spanish word-form using the Levenshtein distance, which measures the edit distance between each pair of phoneme strings [@levenshtein1966binary]. The output indicates the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other. We then divided this value by the length of the longest phoneme string and subtracted the outcome from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different, and one indicates that the two word-forms are identical [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants) for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, multilevel, ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. (1) *Age*, calculated as the number of months elapsed between participants' birth date and questionnaire completion). (2) *Length*, the number of phonemes in the X-SAMPA phonological transcription of the word-form. (3) *Exposure*, a language exposure-weighted lexical frequency. To account for the fact that bilinguals' exposure to a given word-form is a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child who is reportedly exposure to Catalan 80% of the time, and to Spanish 20% of the time, the expected exposure score to the Catalan word *cavall* [*horse*]---with a lexical frequency of `r round(items$freq[items$item=="cat_cavall"], 2)`---would be `r round(items$freq[items$item=="cat_cavall"]*0.80, 2)`, while that of its translation to Spanish *caballo* would be `r round(items$freq[items$item=="cat_cavall"]*0.20, 2)`. (4) *Cognateness*, defined as the phonological similarity between translation equivalents, and calculated as detailed in the *Questionnaire* section.  These predictors were included as main effects in the model, together with their two-way and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data (see @eq-linear) [@barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

@@

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome]. We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four MCMC  chains with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

The main effect of $\text{Age}$ showed the strongest association with the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_age_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_age_std"]]$.lower, 2)`, `r round(post_draws_list[["b_age_std"]]$.upper, 2)`]), with all posterior samples falling out of the ROPE. A one-month increment in age increased a maximum of `r round(post_draws_list[["b_age_std"]]$.median/sd(responses$age), 2)` the probability of acquisition. Similarly, the word exposure index ($\text{Exposure}$) had a strong effect on the probability of acquisition ($\beta$ = `r round(post_draws_list[["b_exposure_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_exposure_std"]]$.lower, 2)`, `r round(post_draws_list[["b_exposure_std"]]$.upper, 2)`]). All of the posterior samples of this regression coefficient excluded the ROPE. The impact of this predictor on the probability of acquisition was positive: for every standard deviation increase in exposure, the participant was `r round(post_draws_list[["b_exposure_std"]]$.median/sd(responses$exposure), 2)` more likely to acquire it. Word-form length also showed a significant association with probability of acquisition ($\beta$ = `r round(post_draws_list[["b_n_phon_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_n_phon_std"]]$.lower, 2)`, `r round(post_draws_list[["b_n_phon_std"]]$.upper, 2)`]). For every phoneme in the word-form, participants were `r round((post_draws_list[["b_n_phon_std"]]$.median)/sd(responses$n_phon), 2)` less likely to know it. The 95% HDI of the regression coefficient of the $\text{Age} \times \text{Exposure}$ interaction also excluded the ROPE ($\beta$ = `r round(post_draws_list[["b_age_std:exposure_std"]]$.median, 2)`, 95% HDI = [`r round(post_draws_list[["b_age_std:exposure_std"]]$.lower, 2)`, `r round(post_draws_list[["b_age_std:exposure_std"]]$.upper, 2)`]), showing that the effect of the word exposure index differed across ages: older children were more sensitive likely to acquire words with higher exposure rate than younger children.

@@

```{r fig-marginal}
#| label: fig-marginal
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: 100%
#| eval: true
#| fig-cap: "Posterior marginal effects. Lines correspond to 100 posterior-predicted medians. Different colours indicate different levels of cognateness (phonological similarity). Black lines indicate the median of the posterior predictions. Predictions are presented separately for different degrees of word exposure index: little exposure to the word, mean exposure, and high exposure. Predictions for Comprehension are shown on top and predictions for Comprehension and Production are shown on the bottom. In-sample predictions lie inside the grey rectangles."
clrs_reorder <- c("#58508d",
                  "#003f5c",
                  "#122A36", 
                  "#ffa600", 
                  "#ff6361", 
                  "#bc5090")

model_epreds |> 
    mutate(age = (age_std * sd(responses$age)) + mean(responses$age),
           exposure = factor(exposure_std,
                             levels = unique(exposure_std),
                             labels = c("Low exposure (-1 SD)",
                                        "Mean exposure",
                                        "High exposure (+1 SD)"),
                             ordered = TRUE),
           lv = factor(lv_std,
                       levels = unique(lv_std),
                       labels = paste0(c(0, 50, 100), "%"),
                       ordered = TRUE),
           .category = ifelse(.category=="Understands",
                              "Comprehension",
                              "Production")) |> 
    ggplot(aes(age, .value, 
               colour = lv, 
               fill = lv)) +
    facet_grid(.category~exposure) +
    annotate(geom = "rect",
             xmin = min(responses$age),
             xmax = max(responses$age),
             ymin = -Inf,
             ymax = Inf,
             fill = "grey95",
             colour = NA) +
    annotate(geom = "text",
             x = mean(responses$age),
             y = 0,
             hjust = -0.1,
             vjust = 0,
             label = "Observed age range",
             size = 2) +
    geom_vline(xintercept = mean(responses$age), 
               colour = "grey") +
    geom_hline(yintercept = 0.5,
               colour = "grey") +
    geom_line(aes(group = interaction(lv, .draw)),
              linewidth = 2/4,
              alpha = 1/5) +
    stat_summary(geom = "line", fun = mean,
                 linewidth = 2/4,
                 colour = "black",
                 aes(linetype = lv,
                     group = interaction(lv, exposure))) +
    labs(x = "Age (months)",
         y = "Probability of acquisition",
         linetype = "Cognateness (phonological similarity)",
         colour = "Cognateness (phonological similarity)",
         fill = "Cognateness (phonological similarity)") +
    guides(colour = guide_legend(override.aes = list(linewidth = 2,
                                                     alpha = 1)),
           linetype = guide_none()) +
    scale_colour_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_fill_manual(values = clrs_reorder[c(4, 5, 1)]) +
    scale_y_continuous(breaks = seq(0, 1, 0.25),
                       limits = c(0, 1)) +
    scale_x_continuous(breaks = seq(0, 50, 5)) +
    theme(legend.position = "top",
          legend.key.size = unit(1, "cm"),
          legend.key.height = unit(1, "cm"),
          axis.text.x = element_text(size = 9),
          panel.grid = element_blank(),
          panel.border = element_rect(fill = NA, 
                                      colour = "black", 
                                      linewidth = 0.75))
```
